Collaborative Visual Analysis on the Web with RCloud

### 2015-03-31 interviews.tex
working

To evaluate the effectiveness of RCloud, we interviewed 13 current and past
users of RCloud. Of these, 9 are data analysts and 4 build tools for data
analysts and business needs.

Generally, analysts were more annoyed by the cell interface and the
perceived slowness of hte interface than web/tool developers. Analysts
switching to R from another language were more likely to use RCloud for
exploratory data analysis. Web/tool developers were more likely to see the
need for parameterization.

\subsection{Sharing of results}
Sharing of notebooks is the core feature of RCloud, and a popular one. All
active users we spoke to praised this feature.

@@

By default, all notebooks are publicly visible in RCloud, and notebooks can be
found through the notebook tree or search. However, the method for sharing that
users mentioned most is sending links through email. Clint says, ``If my supervisor
wants to see what I've done or QA it, I can just send her a link.''

Besides providing a way to present work to others, notebook sharing can also
provide a starting point. Clint says, ``I think the best part about it is how
easily you can share code. [...] You can find a working example, rather than wearing
out Google and finding questionable examples that may or may not work.'' Raif
notes, if ``some person has done something similar, then you're able to just
edit that, and that's saved a lot of work time for me''

Taylor develops packages for analysts, and uses RCloud ``more for sharing code with
other people, and for doing tutorials for iotools or hmr'', his packages.
``I want people to see how the package works, so I clearly want them
to see the code [...] I kind of write it just like I would write GitHub
Markdown, where you have little code snippets and text, but RCloud lets me
actually run the snippets'' and display the results. He also uses RCloud for
explaining problems with data sources.

Some users who tried out RCloud but are not able to use it for organizational
reasons, already miss the functionality. Shelly likes ``the concept of being
able to create notebooks and share them.'' A ``wiki is not the best way for
communicating results - it's kind of like writing a blog post with very limited
functionality [...] I have to save every picture and post it as an image'' and
``I can't share all of the code because it would just get crowded and wouldn't
look right on a wiki.'' Bart says, ``if I could make a folder on RCloud and have
Python notebooks and also Pig notebooks there, and execute them from RCloud,
that would be much better than my current thing, because that would free me from
manual documentation and version control and also telling people where my code
was. It would be just, hey, go look on RCloud, here's the stuff.''

@@

\subsection{Forking}
The ability to start work where someone else let off, by forking, has proved to
be a popular feature. In fact, almost twice as many (131) users have forked
someone else's notebook as have starred one (75).  Clint says, ``it's one of the
handiest functions, because instead of having to find it, copy, paste it, you
just hit Fork, you rename it, and it's done. It's pretty amazing.''

Although we always intended forking to be used to improve others' code, we we
initially didn't support forking one's own notebooks, and this has proved very
useful. Rick says ``I fork my own notebooks because I'm going off and doing some
other analogous project, so I've got interesting content that I've already done
in a previous analysis that I want to start from and then tweak to match a new
set of data.''

Forking also provides a way for others to troubleshoot when something goes
wrong.  When Horace works with the users of his notebooks, ``I'll teach people
to intercept the result in the middle,'' to ``insert print statements here and
there and check values.''

@@

An extremely common but more problematic use of forking is to change
parameters. Raif: ``I've been forking other people's notebooks [because] I
want to run it on a different part of data, or I want to change some parts,
I don't want to see this column, I want a different column, things like
that.''  Taylor complains that a notebook might say ```This is a report of
the volume of all of our feeds for this month', and someone would want to
look at it for the next month or the previous month, so they'd fork it to
change the month.''

Currently parameters can be added to the URL, but adding user interface
elements to do the same thing requires expertise. The tool-builders see a
need for facilities to make it easier. Jyothi calls it ``web-enabling'' the
notebook: users ``can always fork the notebooks and make changes, but I
feel that if the owner of the notebook web-enables it, [...] it's easier to
run. Instead of forking it, if they can set options, it's probably more
efficient, and also they can [still] fork it if they want to.''

@@

Automatic source control is also a popular feature. Clint says, ``instead of
looking back and saying I've got a billion files here in this subdirectory and I
hope I've got them backed up, if they're on RCloud I know they are.''

@@

Bart points out that the automatic versioning works well for dealing with the
minutiae of web development:
\begin{quote}
I like the fact that it has a built-in editor, so
if you need to fix a typo in a link, or an extra line break or any of this other
nonsense, you can just switch to the edit view, pull up your asset, type
something, it's automatically saved, committed, everything. You don't have to go
back to your source code, change it, commit it to the repo, pull the repo to
your distribution version. 
%% None of that's necessary. A lot of web development is
%% fixing stupid little bugs, and you can do that instantly without a lot of
%% overhead, which is nice.
\end{quote}

On the other hand, saving every change leads to many fine-grained
versions. Rick says that for this reason, the history feature is not all that
helpful: ``I don't need something that keeps track of every mistake I've made or
every direction I've tried.''
%% Tagging these versions with names wouldn't
%% necessarily help because ``namespaces are already crowded, so trying to remember
%% the names of notebooks is hard, much less the names of states within each
%% notebook.''

\subsection{Discovering others' work}
Many users find the search function useful. Raif says ``the fact that we have
all the notebooks there, searchable, saves me from replicating what other people
have done.''

But other users prefer to browse just the notebooks of the experts they
know. Rick says, ``usually I know somebody's notebooks that I want to search
through [...] because I know that the kind of thing I'm looking for is something
more obscure than I'm likely to find in some random person's.''

More selective ways to search will be needed as the volume of notebooks
continues to grow. We explore some ideas in Section~\ref{sec:discussion}.

\subsection{Integrated analysis}
Integrating an analysis language into a web development environment is something
the tool developers really appreciate.  The structure of Horace's visualization
notebook means
\begin{quote}
the other guys who want to do analytics on the data, they can first pull
the data, do the analytics on it, and then feed the viewer the data. So
[...] anyone from the stats group [...] can always insert something in
between. [...] Once you get the data into RCloud, then you have a
dataframe to work with, and then they know how to produce another
dataframe.
\end{quote}

Integrating R also helps Jyothi's application: ``Having an open session
where I can run R commands or R functions without having to invoke an API
or send out a request and then wait for the request to come back [...] is
extremely helpful.''

@@

Although most of the analysts use and appreciate the sharing features,
RCloud is less popular as a tool for exploratory data analysis. Every
analyst with prior R experience still does analysis in another tool and
pastes their code into RCloud for sharing.

This is partly out of familiarity. Taylor asks, ``why would I want to use
RCloud over my current setup? If it's just me, I like my text editor and
terminal. There's nothing that I want that those two don't give me.''

@@

The web interface of RCloud doesn't work for Rick: ``It's nice to have
it saved, but there's this trade-off between it making it easier for me to
present something or to save something, and my ease of typing and
correcting and things in a plain old editor window.''

Kenny also works with text file and command line, rearranging a file so the
good code is at the top of the file. When he's done, all the the right is
code at the top. RCloud does not readily support this workflow, but Kenny
often shares the work by pasting it into an RCloud notebook when he's done.

@@

The interface takes some learning. Clint says at first,
\begin{quote}
%% It took me a couple of weeks of
%% looking at it to become comfortable because 
I saw the cells and it just threw me
for a total loop. I mean, it's a good idea because [...] I can run it all at
once if I want or I can break these into sections for either debugging or
staging purposes. I really like it now, but when I first saw it, [it was] very
confusing.
\end{quote}

Just the difference between a cell and the command prompt can confuse and slow
down users. Rick explains:
\begin{quote}
If I was typing into one of the cells near that top, I had
to think ``I'm editing this cell and then I have to execute it,'' and if was doing
the one at the bottom I could type it but then it would automatically execute
[by pressing enter], but then it became a standard cell and I have to edit
it.
\end{quote}

@@

Much of the time, commands typed at the R command line just serve a transitory
purpose, so having RCloud save commands into cells can be annoying. Kenny notes, ``It saves everything I do like everything is gold, but most
of it is junk not meant to be saved.'' Rick says ``I just want to type in a
couple of quick commands and get some results that are going to tell
me what to do next, and they're not necessarily archival in any sense.''

The material that is not appropriate to save includes ``expressions that allow
me to check that I'm the right track'' (Rick), ``checking out what your data is,
or you make a plot of the data. Things that should not really become part of a
notebook, but things that help you understand your data better'' (Raif).

Users also felt that cells do not capture the right level of granularity: they
either hold too much code or too little. Rick says that RCloud's cell structure
``tempts me to type a big long thing and then run the whole thing, as opposed to
typing a few little pieces [...] on the inner parts and then put
them together to make the big thing'' as he would do on the command line. When
Taylor uses the R command line, he copies and pastes ``like 5-10 lines of code,
so when something breaks, I get an error message on that one line, and I can
up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell,
so the only way to get that same functionality is if every line's in one
cell.''

Cells can also take up substantial vertical space, requiring scrolling. Rick
complains that ``every time I type more stuff, the notebook gets longer and
longer and it's harder to deal with.'' Ivan says that cells' controls and blank
space take extra screen space compared to a straight command line, and he and
KC both complained that when there are long results or plots, it causes the
code to scroll off the screen. Raif explains, ``Imagine I run a cell whose
output is huge. [...] To find the next cell, I have to scroll down and find
where that cell starts. So I'm losing the continuity of my code.''

Some users would like the option to keep results separate from code.
Ivan thinks that RStudio's layout is more helpful because charts
are kept in another pane and stay in one place while doing analysis. Raif says
``cells are really useful'' but ``you want to see your output in a different
window [or] on a separate part of the window''.

The cell structure also can be problematic if some cells take a long time to
run. In Kenny's work, there is often a ``long tail''.  The first cell may take
seconds to run, the next cell minutes, and the last cell 5 hours.  In this
situation, the Run Whole Notebook button is ``dangerous''. When converting
his work to notebooks, Kenny ends up with a lot of comments that say ``this cell
takes a long time to run.''

To avert this danger, some users write code within the cells to
cache results. Kenny ends up ``littering the notebook with little switches that
comment out the slow parts'' and either ``load the object instead'' or ``save
the object to disk''. One of the authors of this paper manually writes cells
that check if a result file exists, and perform the calculation only if it's not
there.

@@

\subsection{A sea of notebooks}
RCloud is starting to be a victim of its own success, as it is becoming
difficult to navigate all the notebooks.  There are over 5200 notebooks on
the research instance, of which more than 500 have been starred and more
than 350 have been forked.

For this reason, Rick doesn't find the notebook tree very useful:
\begin{quote}
I don't necessarily need to see everybody's notebook that uses RCloud [...] Every
time I do something new, I get a new notebook and so now my notebooks are maybe
50 or 60. That's enough to think about just on my own, but if I've got
everybody's 50 or 60 sitting on my display, I find that it's more than I want to
know about.
\end{quote}

Although RCloud promises an environment where notebooks stay working, our
users have not all learned the habits that make this a reality. As Byers
puts it, there is still a big problem with ``bitrot'' -- notebooks often
stop working because the user changed the structure of their data, or
changed a filename or a database.  He says we need ``organizational
protocols'' to catch up with the technology. Even if one forks someone
else's notebook and corrects it, the original notebook still exists with
the error.

@@

Taylor, who writes packages and example notebooks for them, seems to be
haunted the most by dead notebooks. Taylor and Horace work together on a notebook:
\begin{quote}
There's no way for both of us to have ownership of a notebook, so the only
way is to fork it back and forth, and so we have dozens of old copies. [We
end up] deleting all the old ones, but people still have links to them,
because they don't actually disappear.
\end{quote}

The problem of dead notebooks is compounded by changing
packages and their example notebooks at the same time. Taylor will ``go back and
change it and update the page, but then what happens is people have in the
meantime forked it. [...]  I'll have to change a package as well, so their
old fork stops working, and then they complain.''

Taylor tried to keep his notebook tree tidy, but this didn't help, because
people either kept old links in their email, or forked a notebook which is now
obsolete. By design, notebooks that are deleted in the user interface are not
purged from the repository. ``Now I'm really gun-shy about sharing notebooks
because it's like, do I want to support this forever?''

>>>>

To evaluate the effectiveness of RCloud, we interviewed 13 current or
recent users of RCloud. 9 are data analysts, and 4 build tools for data
analysts and business needs.

The subjects differed in their evaluation of its benefits and disadvantages,
as described below.

%Analysts were more annoyed by the cell interface and its perceived
%slowness than were web/tool developers. Analysts
%switching to R from another language were more likely to use RCloud for
%exploratory data analysis. Web/tool developers were more likely to see the
%need for parameterization.

\subsection{Sharing of results}
Sharing of notebooks is the core feature of RCloud, and a popular one. All
the subjects praised this feature.

@@

By default, all RCloud notebooks are publicly visible, and notebooks can be
found by navigating the notebook tree, or by searching. However, users most
often mentioned sharing by sending links through email. Clint says,
``If my supervisor wants to see what I've done or QA it, I can just send her a link.''

Besides providing a way to present work, notebook sharing can provide a
starting point for coding. Clint says, ``The best part is how easily you can
share code. You can find a working example, rather than wearing
out Google and finding questionable examples that may or may not work.'' Raif
notes, ``[If] some person has done something similar, then you're able to just
edit that, and that's saved a lot of work time for me.''

Taylor develops packages for analysts, and uses RCloud ``more for sharing code with
other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
``I want people to see how the package works, so I clearly want them
to see the code... I write it just like I would write GitHub
Markdown, where you have little code snippets and text, but RCloud lets me
actually run the snippets [and display the results].'' He also uses RCloud for
describing and debugging  problems with data sources.

Some users, who tried RCloud and were not able to continue for organizational
reasons, miss certain capabilities. Shelly ``[likes] the concept of being
able to create notebooks and share them... A wiki is not the best way for
communicating results - it's kind of like writing a blog post with very limited
functionality. I have to save every picture and post it as an image.'' She said,
``I can't share all of the code because it would just get crowded and wouldn't
look right on a wiki.'' Bart explained, ``If I could make a folder on RCloud and
have Python notebooks and also Pig notebooks there, and execute them from RCloud,
that would be much better than my current [environment], because that would free
me from manual documentation and version control and also telling people where
my code was. It would be just, hey, go look on RCloud, here's the stuff.''

@@

\subsection{Forking}
The ability to start work where someone else let off, by forking, proved to
be a popular feature. In fact, almost twice as many (131) users have forked
someone else's notebook as have starred one (75).  Clint says, ``It's one of the
handiest functions, because instead of having to find it, copy, paste it, you
just hit Fork, rename it, and it's done. It's pretty amazing.''

Although we intended forking to be available to improve others' code, we 
initially didn't anticipate support forking one's own notebooks, which
proved very useful. Rick says, ``I fork my own notebooks because I'm going off
and doing some other analogous project, so I've got interesting content that I've
already done in a previous analysis, that I want to start from and then tweak
to match a new set of data.''

Forking also provides a way for others to troubleshoot when something goes
wrong.  When Horace works with the users of his notebooks, ``I'll teach people
to intercept the result in the middle, to insert print statements here and
there and check values.''

@@

A common but somewhat problematic use of forking is to change
parameters. Raif: ``I've been forking other people's notebooks [because] I
want to run them on a different part of data, or I want to change some parts,
I don't want to see this column, I want a different column, things like
that.''  Taylor complains that a notebook might say ```This is a report of
the volume of all of our feeds for this month', and someone would want to
look at it for the next month or the previous month, so they'd fork it to
change the month.''

Parameters can be added to a notebook URL, but adding user interface
elements to do the same thing requires expertise. Tool builders see a
need for facilities to make it easier. Jyothi calls it ``web-enabling'' the
notebook: ``[users] can always fork the notebooks and make changes, but I
feel that if the owner of the notebook web-enables it, it's easier to
run. Instead of forking it, if they can set options, it's probably more
efficient, and they can [still] fork it if they want to.''

@@

Automatic source control is also a popular feature. Clint says, ``Instead of
looking back and saying I've got a billion files here in this subdirectory and I
hope I've got them backed up, if they're on RCloud I know they are.''

@@

Bart points out that the automatic versioning works well for dealing with the
minutiae of web development:
\begin{quote}
I like the fact that it has a built-in editor, so
if you need to fix a typo in a link, or an extra line break or other
nonsense, you can just switch to the edit view, pull up your asset, type
something, it's automatically saved, committed, everything. You don't have to go
back to your source code, change it, commit it to the repo, pull the repo to
your distribution version. 
%% None of that's necessary. A lot of web development is
%% fixing stupid little bugs, and you can do that instantly without a lot of
%% overhead, which is nice.
\end{quote}

On the other hand, saving every change leads to many fine-grained
versions. Rick says that for this reason, the history feature is not that
helpful: ``I don't need something that keeps track of every mistake I've made or
every direction I've tried.''
%% Tagging these versions with names wouldn't
%% necessarily help because ``namespaces are already crowded, so trying to remember
%% the names of notebooks is hard, much less the names of states within each
%% notebook.''

\subsection{Discovering others' work}
Many users find the search function valuable. Raif says, ``The fact that we have
all the notebooks there, searchable, saves me from replicating what other people
have done.''

But other users prefer to browse just the notebooks of experts they
know. Rick says, ``Usually I know somebody's notebooks that I want to search
through, because I know that the kind of thing I'm looking for is something
more obscure than I'm likely to find in some random person's.''

More selective ways to search will be needed as the number of notebooks
grows further. We explore some ideas in Section~\ref{sec:discussion}.

\subsection{Integrated analysis}
Integrating an analysis language into a web development environment is something
tool developers really appreciate.  The structure of Horace's visualization
notebook means
\begin{quote}
the other guys who want to do analytics on the data can first pull
the data, do the analytics on it, and then feed the viewer the data.
Anyone from the stats group can insert something in between.
Once you get the data into RCloud, then you have a dataframe
to work with, and then they can produce another dataframe.
\end{quote}

Integrating R also helps Jyothi's topic modeling tool, described in the
last section: ``Having an open session where I can run R commands or
functions without having to invoke an API or send a request and then
wait for the response is extremely helpful in writing the application.''

@@

Although most of the analysts use and appreciate the sharing features,
RCloud is less popular as a tool for exploratory data analysis. Every
analyst with prior R experience still prefers to do analysis in another
tool, and paste code into RCloud for sharing.

This is mainly out of familiarity. Taylor asks, ``Why would I want to use
RCloud over my current setup? If it's just me, I like my text editor and
terminal. There's nothing that I want that those two don't give me.''

@@

The web interface of RCloud isn't satisfactory to Rick: ``It's nice to have
it saved, but there's this trade-off between it making it easier for me to
present something or to save something, and my ease of typing and
correcting and things in a plain editor window.''

Kenny also works with text files and command line, rearranging a file so the
good code is at the top of the file. When he's done, all the right code is
at the top. RCloud does not readily support this workflow, but Kenny
often shares work by pasting it into an RCloud notebook when he's done.

@@

The interface takes some learning. Clint says,
\begin{quote}
%% It took me a couple of weeks of
%% looking at it to become comfortable because 
I saw the cells and it just threw me
for a total loop. I mean, it's a good idea because [...] I can run it all at
once if I want or I can break these into sections for either debugging or
staging purposes. I really like it now, but when I first saw it, [it was] very
confusing.
\end{quote}

Simply the difference between a cell and the command prompt can confuse some people.
Rick explains:
\begin{quote}
If I was typing into one of the cells near that top, I had to think ``I'm editing
this cell and then I have to execute it,'' and if I was editing one at the bottom,
I could type it but then it would automatically execute [by pressing enter].
%, then it became a standard cell and I have to edit it.
\end{quote}

@@

Much of the time, commands typed at the R command line serve only a transitory
purpose, so having RCloud persist commands in cells can be annoying.
Kenny notes, ``It saves everything I do like everything is gold, but most
of it is junk not meant to be saved.'' Rick says, ``I just want to type in a
couple of quick commands and get some results that are going to tell
me what to do next, and they're not necessarily archival in any sense.''

Material that is not appropriate to save includes ``expressions that allow
me to check that I'm the right track'' (Rick), ``checking out what your data is,
or you make a plot of the data. Things that should not really become part of a
notebook, but things that help you understand your data better'' (Raif).

Users felt that cells do not capture the right level of granularity: they
either hold too much code or too little. Rick says that RCloud's cell structure
``tempts me to type a big long thing and then run the whole thing, as opposed to
typing a few little pieces and then put them together'' as he would on the command line.
When Taylor uses the R command line, he ``copies and pastes 5-10 lines of code,
so when something breaks, I get an error message on that one line, and I can
up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell,
so the only way to get that functionality is if every line's in one cell.''
Another complaint related to the human interface was that cells can take up
substantial vertical space, requiring scrolling.
% Rick complains, ``Every time I type more stuff, the notebook gets longer and
% longer and it's harder to deal with.'' Ivan says that cells' controls and blank
% space take extra screen space compared to a straight command line, and he and
% KC both complained that when there are long results or plots, it causes the
% code to scroll off the screen. Raif explains, ``Imagine I run a cell whose
% output is huge. [...] To find the next cell, I have to scroll down and find
% where that cell starts. So I'm losing the continuity of my code.''

Some users would like a way to keep results separate from code.
Ivan thinks RStudio's layout is more helpful because charts
are shown in another pane that stays in place while doing analysis. Raif says
``Cells are really useful, [but] you want to see your output in a different
window or on a separate part of the window''.

The cell structure also can be problematic if some cells take a long time to
execute. In Kenny's work, there is often a ``long tail''.  The first cell may
take just seconds to run, the next cell minutes, and the last cell 5 hours.
In this situation, the Run Whole Notebook button is dangerous.
Kenny reports that when converting his work to notebooks, he
ends up with a lot of comments saying ``This cell takes a long time to run.''

To avert this pitfall, some users write code within cells to explicitly cache results.
Kenny ends up ``littering the notebook with little switches that comment out the
slow parts'' and either load or save the object to disk. One of the authors of this paper
manually writes cells that check if a result file exists, and perform a calculation only
if it's needed.

@@

\subsection{A sea of notebooks}
RCloud is becoming a victim of its own success, as it is becoming
difficult to navigate all the notebooks.  There are over 5200 notebooks in
the research instance, of which more than 500 have been starred, and more
than 350 have been forked.

For this reason, Rick doesn't find the notebook tree satisfactory:
\begin{quote}
I don't necessarily need to see everybody's notebook that uses RCloud. Every
time I do something new, I get a new notebook, so now I have 50 or 60 notebooks.
That's enough to think about just on my own, but if everybody has
50 or 60 sitting on my display, it's more than I want to know about.
\end{quote}

Although RCloud promises an environment where notebooks should keep working,
not all our users have learned the habits that make this a reality. As Byers
puts it, there is still a big problem with ``bitrot'' -- notebooks often
stop working because the user changed the structure of their data, or
changed a filename or a database.  He says we need ``organizational
protocols'' to catch up with the technology. Even if one forks someone
else's notebook and corrects it, the original notebook still exists with
the error.

@@

Taylor, who writes packages and example notebooks for them, mentioned the
accumulation of dead notebooks. When Taylor and Horace work together on a notebook:
\begin{quote}
There's no way for both of us to have ownership of a notebook, so the only
way is to fork it back and forth, and so we have dozens of old copies. We
end up deleting all the old ones, but people still have links to them,
because they don't actually disappear.
\end{quote}

The problem of dead notebooks is compounded by changing
packages and example notebooks at the same time. Taylor will ``go back and
change it and update the page, but then what happens is people have in the
meantime forked it. I'll have to change a package as well, so their
old fork stops working, and they complain.''

Taylor tried to keep his notebook tree well organized, but this didn't help,
because some people kept old links in their email, or forked notebooks which 
had become obsolete. By design, notebooks that are deleted in the user interface
are not purged from the repository. This caused some subjects to express
caution about sharing notebooks too freely.
%``Now I'm really gun-shy about sharing notebooks
%because it's like, do I want to support this forever?''
